### 2013-08-23 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
more edits to the nonlinear chapter

\subsection{Automatic Differentiation for Jacobian Generation}
\label{subsec:automatic_differentiation}
If it is acceptable to store the actual Jacobian matrix, other methods
are available to construct it without requiring hand-coding and
evaluating derivatives, thus eliminating the associated issues. In
addition, if any additional equations are added to the system or a
higher order functional approximation is desired, it would be useful
to avoid regenerating and coding these derivatives. Becoming more
prominent in the 1990's, \textit{automatic differentiation} is a
mechanism by which the derivatives of a function can be generated
automatically by evaluating it. Automatic differentiation is built on
the concept that all functions discretely represented in a computer
are ultimately represented by elementary mathematical operations. If
the chain rule is applied to those elementary operations, then the
derivatives of those functions can be computed to the order of
accuracy of their original discretization in a completely automated
way \cite{averick_computing_1994}.

@@

%%---------------------------------------------------------------------------%%
\section{The FANM Method\ }
\label{sec:fanm}
In physics codes based on nonlinear equations sets, Newton-Krylov
methods are the primary means of generating a fully consistent
solution scheme
\cite{evans_development_2006,evans_enhanced_2007,gaston_parallel_2009,godoy_parallel_2012}. It
is common that for large scale simulations these problems are memory
limited due to the subspaces generated by robust Krylov methods which
may often build hundreds of vectors during a Newton step. Often, a
matrix-free approach is chosen to relax memory requirements over
directly generating the Jacobian matrix and facilitate the
implementation. However, as we observed in previous sections, these
matrix-free methods suffer from poorly scaled problems and the first
order error introduced by the Jacobian approximation. In addition, it
was observed that the savings induced by the matrix-free approach is
eventually amortized over a number of nonlinear iterations where it
becomes more efficient computationally to instead form the Jacobian.

@@

\subsection{Thermal Convection Cavity Problem}
\label{subsec:natural_convection_cavity}
In 1983 a set of benchmark solutions for the natural convection of air
in a square cavity was published \cite{de_vahl_davis_natural_1983} as
shown in Figure~\ref{fig:natural_convection_cavity} for the solution
of the energy, mass, and momentum equations.

\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/nonlinear_problem/natural_convection_cavity.pdftex_t} }
  \end{center}
  \caption{\textbf{Problem setup for the natural convection cavity
      benchmark.} \textit{Dirichlet conditions are set for the
      temperature on the left and right while Neumann conditions are
      set on the top and bottom of the Cartesian grid. The temperature
      gradients will cause buoyancy-driven flow. Zero velocity
      Dirichlet conditions are set on each boundary. No thermal source
      was present.}}
  \label{fig:natural_convection_cavity}
\end{figure}

In this problem, a rectilinear grid is applied to the unit square. No
heat flow is allowed out of the top and bottom of the square with a
zero Neumann condition specified. Buoyancy driven flow is generated by
the temperature gradient from the hot and cold Dirichlet conditions on
the left and right boundaries of the box. By adjusting the Rayleigh
number of the fluid (and therefore adjusting the ratio of convective
to conductive heat transfer), we can adjust the influence of the
nonlinear convection term in Eq~(\ref{eq:ns_momentum}). In Shadid's
work, Rayleigh numbers of up to \sn{1}{6} were used for this benchmark.

@@

In this problem, the inflow condition is specified by a fully-formed
parabolic flow profile over a zero velocity boundary representing a
step. For this work, a slightly different parabolic formulation for
the inlet velocities was used than in the benchmark. In the benchmark
solutions, the average velocity in the parabola was 2/3 that of the
maximum velocity while the formula given in
Figure~\ref{fig:backward_facing_step} produces a more plug-like flow
with an average velocity 3/4 that of the maximum velocity. The flow
over this step will generate a recirculating backflow under the inlet
flow towards the step and at higher Reynolds numbers a recirculating
zone will form farther downstream from the step. As in the lid driven
cavity problem, the nonlinear behavior of this benchmark and the
difficulty in obtaining a solution is dictated by the Reynolds number
of the fluid. In Shadid's work, a Reynolds numbers up to 800 were
used.

Compared to the lid driven cavity benchmark, the flow over a backward
step is much more representative of flow driven by inertial forces in
a nuclear reactor that may encounter several obstacles along the flow
path. For example, spacer grids in fuel assemblies or mixing vanes
designed to induce turbulence and improve heat transfer will create
recirculation zones much like those observed for this benchmark
problem. Figure~\ref{fig:step_velocity_isocurves} gives the velocity
magnitude contours for solutions to this problem with Reynolds numbers
of 200, 300, 400, and 500.

@@

For each of the benchmark problems, we will present the results of
computations using FANM and compare them to results using a
Newton-Krylov method in order to verify the correctness of FANM and
its applicability to fluid problems. Given the difficulty of these
benchmark problems, the linear solve at each Newton step was
preconditioned with an algebraic multigrid method\footnote{The ML
  algebraic multigrid library in the Trilinos scientific computing
  libraries \cite{heroux_overview_2005} was used for all problems.}, a
common technique for preconditioning these equations
\cite{ghia_high-re_1982,evans_enhanced_2007} with the preconditioner
reconstructed at every Newton step for the new Jacobian. Without
multigrid preconditioning, the residual in MCSA was observed to
stagnate. Typically this happens because the Richardson iteration and
Neumann series on which the Monte Carlo method is based will quickly
damp the higher order error modes on the scale of the computational
grid used to solve the problem but will reduce lower order error modes
at a considerably lower rate. In terms of Monte Carlo, this stagnation
manifests itself in histories that simply do no terminate and
therefore no solution is acquired. For many of the problems presented
here, it is likely that a more physics-based preconditioning strategy
such as those presented in
\cite{evans_development_2006,evans_enhanced_2007} will yield better
results for more difficult problems. However, the explicit
preconditioning strategy used here currently limits us to algebraic
preconditioning.

For every benchmark calculation, both FANM and Newton-Krylov were
preconditioned with the same multigrid parameters such that the
conditioning of the system would not be a factor in the verification
of the method or the performance comparison in the following
section. With FANM, this preconditioning was required to reduce the
spectral radius of the Jacobian operators generated at each Newton
step. This preconditioning was also required to obtain good
convergence properties for the Newton-Krylov solutions although it was
not necessary for their convergence. In addition, the same nonlinear
solver parameters including globalization and forcing term strategies
were used with both methods with a nonlinear convergence criteria of a
\sn{1}{-12} norm for the nonlinear residual. As with the radiation
transport calculations in the previous chapter, algebraic
preconditioning with the multigrid method results in dense linear
systems. To mitigate this, the reduced domain approximation was again
applied to the Monte Carlo solution within the MCSA solve occurring at
each nonlinear iteration. The values used for the approximation will
be given for each benchmark problem.

Results in this and the following section were produced by the Drekar
multiphysics code base being developed at Sandia National Laboratories
for fluid flow and other multiple physics simulations
\cite{pawlowski_drekar_2012}. Drekar is a massively parallel physics
framework that utilizes Newton-Krylov methods to generate a fully
consistent solution scheme for the Navier-Stokes equations among other
systems with Jacobian generation through automatic differentiation. To
achieve these results, a Monte Carlo solver generated for this work
was injected into the nonlinear tool chain within Drekar to form an
implementation of the FANM method. For the Newton-Krylov solutions,
the same implementation of GMRES from the Trilinos scientific
computing library Aztec \cite{heroux_overview_2005} used for the
verification of MCSA for the $SP_N$ solutions in the previous chapter
was also used here as a means of comparison and FANM benchmarking.

@@

For the verification calculations, at each Rayleigh number the fluid
velocity in the x and y directions, fluid pressure, and fluid
temperature values calculated Newton-Krylov were compared against the
Newton-Krylov calculations. No variation in solution was noted for
each of the computations larger than the floating point tolerance
reported by Drekar with an absolute difference value of zero observed
at all spatial locations for each quantity.

@@

For the verification calculations, at each Reynolds number the fluid
velocity in the x and y directions and fluid pressure values
calculated Newton-Krylov were compared against the Newton-Krylov
calculations. No variation in solution was noted for each of the
computations larger than the floating point tolerance reported by
Drekar with an absolute difference value of zero observed at all
spatial locations for each quantity.

@@

Using the benchmarks from the verification, we now compare the
performance of the FANM method against a conventional Newton-Krylov
(NK) method. For each benchmark, we will analyze the iterative
performance of both the nonlinear Newton solver and the linear solver
used to compute the Newton correction vector at each step. Limiting
the number of nonlinear iterations required for solution is critical
for efficiency as nonlinear function evaluations, Jacobian generation,
and preconditioner construction are typically required at every Newton
iteration. In addition, timing results will also be discussed. To
quantify the acceleration provided by the residual Monte Carlo
component of MCSA, each calculation was also performed using the same
Newton method but with a plain preconditioned Richardson iteration as
the linear solution technique to give a Newton-Richardson (NR)
scheme. For all cases the Richardson iteration was preconditioned
in the same way as GMRES and MCSA.

>>>>

\subsection{Automatic Differentiation for Jacobian Generation}
\label{subsec:automatic_differentiation}
If it is acceptable to store the actual Jacobian matrix, methods are
available to construct it without requiring hand-coding and evaluating
derivatives, thus eliminating the associated issues. In addition, if
any additional equations are added to the system or a higher order
functional approximation is desired, it would be useful to avoid
regenerating and coding these derivatives. Becoming more prominent in
the 1990's, \textit{automatic differentiation} is a mechanism by which
the derivatives of a function can be generated automatically by
evaluating it. Automatic differentiation is built on the concept that
all functions discretely represented in a computer are ultimately
represented by elementary mathematical operations. If the chain rule
is applied to those elementary operations, then the derivatives of
those functions can be computed to the order of accuracy of their
original discretization in a completely automated way
\cite{averick_computing_1994}.

@@

%%---------------------------------------------------------------------------%%
\section{The FANM Method\ }
\label{sec:fanm}
In physics codes based on nonlinear equations sets, Newton-Krylov
methods are the primary means of generating a fully consistent
solution scheme
\cite{evans_development_2006,evans_enhanced_2007,gaston_parallel_2009,godoy_parallel_2012}. It
is common that for large scale simulations these problems are memory
limited due to the subspaces generated by robust Krylov methods which
may often build hundreds of vectors during a Newton step. Often, a
matrix-free approach is chosen to relax memory requirements over
directly generating the Jacobian matrix and to facilitate the
implementation. However, these matrix-free methods suffer from the
first order error introduced by the Jacobian approximation in poorly
scaled problems \citep{kelley_iterative_1995}. In addition, it has
been observed that the savings induced by the matrix-free approach is
eventually amortized over a number of nonlinear iterations where it
becomes more efficient computationally to instead form the Jacobian
\citep{knoll_newton-krylov_1995}.

@@

\subsection{Thermal Convection Cavity Problem}
\label{subsec:natural_convection_cavity}
In 1983 a set of benchmark solutions for the natural convection of air
in a square cavity was published \cite{de_vahl_davis_natural_1983} as
shown in Figure~\ref{fig:natural_convection_cavity} for the solution
of the energy, mass, and momentum equations.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/nonlinear_problem/natural_convection_cavity.pdftex_t} }
  \end{center}
  \caption{\textbf{Problem setup for the natural convection cavity
      benchmark.} \textit{Dirichlet conditions are set for the
      temperature on the left and right while Neumann conditions are
      set on the top and bottom of the Cartesian grid. The temperature
      gradients will cause buoyancy-driven flow. Zero velocity
      Dirichlet conditions are set on each boundary. No thermal source
      was present.}}
  \label{fig:natural_convection_cavity}
\end{figure}
In this problem, a rectilinear grid is applied to the unit square. No
heat flow is allowed out of the top and bottom of the square with a
zero Neumann condition specified. Buoyancy driven flow is generated by
the temperature gradient from the hot and cold Dirichlet conditions on
the left and right boundaries of the box. By adjusting the Rayleigh
number of the fluid (and therefore adjusting the ratio of convective
to conductive heat transfer), we can adjust the influence of the
nonlinear terms in Eq~(\ref{eq:ns_momentum}) and
Eq~(\ref{eq:ns_energy}). In Shadid's work, Rayleigh numbers of up to
\sn{1}{6} were used for this benchmark.

@@

In this problem, the inflow condition is specified by a fully-formed
parabolic flow profile over a zero velocity boundary representing a
step. For this work, a slightly different parabolic formulation for
the inlet velocities was used than in the benchmark. In the benchmark
solutions, the average velocity in the parabola was 2/3 that of the
maximum velocity while the formula given in
Figure~\ref{fig:backward_facing_step} produces a more plug-like flow
with an average velocity 3/4 that of the maximum velocity. The flow
over this step will generate a recirculating backflow under the inlet
flow towards the step and at higher Reynolds numbers a recirculating
zone will form farther downstream from the step. As in the lid driven
cavity problem, the nonlinear behavior of this benchmark and the
difficulty in obtaining a solution is dictated by the Reynolds number
of the fluid.

Figure~\ref{fig:step_velocity_isocurves} gives the velocity magnitude
contours for solutions to this problem with Reynolds numbers of 200,
300, 400, and 500. As the Reynolds number increases, note the
downstream recirculation zone beginning to form in the upper right
corner of the solution plots where the velocity isocurves are
depressed. In Shadid's work, Reynolds numbers up to 800 were used.

@@

For each of the benchmark problems, we will present the results of
computations using FANM and compare them to results using a
Newton-Krylov method in order to verify the correctness of FANM and
its applicability to fluid problems. Given the difficulty of these
benchmark problems, the linear solve at each Newton step was
preconditioned with an algebraic multigrid method\footnote{The ML
  algebraic multigrid library in the Trilinos scientific computing
  libraries \cite{heroux_overview_2005} was used for all problems.}, a
common technique for preconditioning these equations
\cite{ghia_high-re_1982,evans_enhanced_2007}. The preconditioner is
reconstructed at every Newton step for the new Jacobian and used on
the right as outlined in
\S~\ref{subsubsec:general_mcsa_preconditioning}. Without multigrid
preconditioning, the residual in MCSA was observed to
stagnate. Typically this happens because the Richardson iteration and
subsequently the Neumann series will quickly damp the higher order
error modes on the scale of the computational grid used to solve the
problem but will reduce lower order error modes at a considerably
lower rate. In terms of Monte Carlo, this stagnation manifests itself
in histories that simply do no terminate and therefore no solution is
acquired. For many of the problems presented here, it is likely that a
more physics-based preconditioning strategy such as those presented in
\cite{evans_development_2006,evans_enhanced_2007} will yield better
results for more difficult problems. However, in this work we will
restrict our attention to multigrid for simplicity.

For every benchmark calculation, both FANM and Newton-Krylov were
preconditioned with the same multigrid parameters such that the
conditioning of the system would not be a factor in the verification
of the method or the performance comparison in the following
section. With FANM, this preconditioning was required to reduce the
spectral radius of the Jacobian operators generated at each Newton
step. This preconditioning was also required to obtain good
convergence properties for the Newton-Krylov solutions although it was
not necessary for their convergence. In addition, the same nonlinear
solver parameters including globalization and forcing term strategies
were used for both methods with a \sn{1}{-12} norm for the nonlinear
residual set as the nonlinear convergence criteria. As with the
radiation transport calculations in the previous chapter, algebraic
preconditioning with the multigrid method results in dense linear
systems. To mitigate this, the reduced domain approximation was again
applied to the Monte Carlo solution within the MCSA solve occurring at
each nonlinear iteration. The values used for the approximation will
be given for each benchmark problem.

Results in this and the following section were produced by the Drekar
multiphysics code base being developed at Sandia National Laboratories
for fluid flow and other multiple physics simulations
\cite{pawlowski_drekar_2012}. Drekar is a massively parallel physics
framework that utilizes a finite element formulation of the
Navier-Stokes equations and Newton-Krylov methods with Jacobian
generation through automatic differentiation. To achieve these
results, a Monte Carlo solver generated for this work was injected
into the nonlinear solver sequence within Drekar to form an
implementation of the FANM method. For the Newton-Krylov solutions,
the same implementation of GMRES used for the $SP_N$ verification from
the Trilinos scientific computing library Aztec
\cite{heroux_overview_2005} was also used here as a means of
comparison and FANM benchmarking.

@@

At each Rayleigh number the fluid velocity in the x and y directions,
fluid pressure, and fluid temperature values calculated with FANM were
verified by computing the point-wise absolute difference with the
Newton-Krylov results. No variation in solution was noted for each of
the computations larger than the floating point tolerance reported by
Drekar with an absolute difference value of zero observed at all
spatial locations for each quantity.

@@

At each Reynolds number the fluid velocity in the x and y directions
and fluid pressure values calculated with FANM were verified by
computing the point-wise absolute difference with the Newton-Krylov
results. No variation in solution was noted for each of the
computations larger than the floating point tolerance reported by
Drekar with an absolute difference value of zero observed at all
spatial locations for each quantity.

@@

Using the benchmarks from the verification, we now compare the
performance of the FANM method against a conventional Newton-Krylov
(NK) method. For each benchmark, we will analyze the iterative
performance of both the Newton solver and the linear solver used to
compute the Newton correction vector at each step. Limiting the number
of nonlinear iterations required for solution is critical for
efficiency as nonlinear function evaluations, Jacobian generation, and
preconditioner construction are typically required at every Newton
iteration. In addition, timing results will also be discussed. To
quantify the acceleration provided by the residual Monte Carlo
component of MCSA, each calculation was also performed using the same
Newton method but with a preconditioned Richardson iteration as the
linear solution technique to give a Newton-Richardson (NR) scheme. For
all cases the Richardson iteration was preconditioned with multigrid
in the same way as GMRES and MCSA.
